Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru

### 2020-08-10 text/ecography/round_3/supplementary_information.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: no
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}
  # Ecography requires double line spacing and line numbering
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers

title: Supplementary Information
bibliography: transect_refs.bib
csl: ecography.csl
# biblio-style: "natbib"
link-citations: true
---

@@

```{r sampling-zones, echo=FALSE, fig.cap="Study area stratified by distance to sea in 10 classes."}
p_1 = readRDS(here("images/18_sample_stratification.rds"))
print(p_1)
```

@@

We established 12 experimental plots of 3 × 3 m^2 in a sandy area typical of the region.
Plots were located on the campus of the Universidad de Piura (Sechura Desert; 5°10'S, 80°38'W) on a sandy plain devoid of woody vegetation.
We built a barbed wire fence around the experimental area to avoid the intrusion of domestic animals such as deer wandering the campus.
The soils of the region are nutrient-poor (e.g., mean nitrogen content: 0.005 mg g^-1^) with high water infiltration rates [@muenchow_woody_2013].
The vegetation surrounding the experimental area mirrors the typical plant formation around Piura, hence a transition between tropical desert and open-shrubland vegetation [@muenchow_coupling_2013].
Prior to the irrigation experiment, we clipped the scarce aboveground biomass.
Between December 2012 and May 2013 four plots received the simulated rainfall amount of the Super El Niño event in 1997/98 (1780 mm), another four received the rainfall amount of the moderate El Niño event in 1991/92 (258 l mm) and the remaining four plots served as baseline plots without any additional water input.
Natural rainfall amounted to 53.9 mm during the study period (data: meteorological station of the University of Piura).
If natural rainfall occurred, it was subtracted from the artificial water input in the case of the Normal and Super Niño plots.
The irrigation water stemmed from rain-collecting tanks.
Irrigation was applied using hand sprinklers [@ronnenberg_effects_2011] before and after sunset and followed the intensity and frequency of the rainfall events of the El Niño episodes in 1991/92 and 1997/98 as recorded by the automatic climatic station of the University of Piura.
On the one hand, this reproduced the temporal pattern of El Niño rainfalls, which commonly took place at night time.
An on the other hand, this procedure impeded undesired evaporation and subsequent salinization effects.
The plan sandy underground prevented any runoff from irrigated plots to adjacent plots.

@@

\subsection*{Appendix 3. Number of species plotted against time during the duration of the experiment.}

```{r alpha-exp, echo=FALSE, fig.cap="Number of species plotted against time during the duration of the experiment.", message=FALSE}
p_1 = readRDS(file = here("images/18_irrigation_spri.rds"))
print(p_1)
```

>>>>

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}
  # Ecography requires double line spacing and line numbering
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  # Use Appendix instead of Figure in the caption
  - \renewcommand{\figurename}{Appendix}

title: Supplementary Information
bibliography: ../transect_refs.bib
csl: ../ecography.csl
# biblio-style: "natbib"
link-citations: true
---

@@

```{r sampling-zones, echo=FALSE, fig.align="center",fig.cap="Study area stratified by distance to sea in 10 classes."}
p_1 = readRDS(here("images/18_sample_stratification.rds"))
print(p_1)
```

@@

We established 12 experimental plots of 3 × 3 m^2^ in a sandy area typical of the region.
Plots were located on the campus of the Universidad de Piura (Sechura Desert; 5°10'S, 80°38'W) on a sandy plain devoid of woody vegetation.
We built a barbed wire fence around the experimental area to avoid the intrusion of domestic animals such as deer wandering the campus.
The soils of the region are nutrient-poor (e.g., mean nitrogen content: 0.005 mg g^-1^) with high water infiltration rates [@muenchow_woody_2013].
The vegetation surrounding the experimental area mirrors the typical plant formation around Piura, hence a transition between tropical desert and open-shrubland vegetation [@muenchow_coupling_2013].
Prior to the irrigation experiment, we clipped the scarce aboveground biomass.
Between December 2012 and May 2013 four plots received the simulated rainfall amount of the Super El Niño event in 1997/98 (1780 mm), another four received the rainfall amount of the moderate El Niño event in 1991/92 (258 l mm) and the remaining four plots served as baseline plots without any additional water input.
Natural rainfall amounted to 53.9 mm during the study period (data: meteorological station of the University of Piura).
If natural rainfall occurred, it was subtracted from the artificial water input in the case of the Normal and Super Niño plots.
The irrigation water stemmed from rain-collecting tanks.
Irrigation was applied using hand sprinklers [@ronnenberg_effects_2011] before and after sunset and followed the intensity and frequency of the rainfall events of the El Niño episodes in 1991/92 and 1997/98 as recorded by the automatic climatic station of the University of Piura.
On the one hand, this reproduced the temporal pattern of El Niño rainfalls, which commonly took place at night time.
An on the other hand, this procedure impeded undesired evaporation and subsequent salinization effects.
The plan sandy underground prevented any runoff from irrigated plots to adjacent plots.

@@

\subsection*{Appendix 3. Relationship between the response and NDVI by year.}

<!-- since we have no Figure in Appendix 2, we have to increase the figure counter by 1, then our spatial prediction will correctly displayed as Appendix 3-->
\setcounter{figure}{2}

```{r dca-ndvi, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between the response (DCA1) and NDVI by year. To aid the visual interpretation, line smoothers were added.", out.width="70%"}
data = readRDS(here("images/15_model_input_data.rds"))
ggplot(data, aes(x = ndvi, y = DCA1, group = year)) +
  geom_point(aes(color = as.factor(year)), alpha = 0.4) + 
  labs(color = "year") +
  xlab(label = "NDVI") +
  ylab(label = "DCA1") +
  geom_smooth(aes(color = as.factor(year)), se = FALSE) + 
  theme_bw() +
  coord_fixed(ratio = 0.15)
```

\subsection*{Appendix 4. Predictive mapping of species richness.}

```{r spri-dist, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, warning=FALSE, fig.cap="Spatial prediction of species richness along the cross-section for different ENSO episodes using the NDVI and year interaction as predictors (used model: GAM with a spline smoother with 3 degrees of freedom and a Poisson distribution). The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour."}
library("sp")
library("sf")
load(here("figures/15_spri_preds.rda"))
print(spri_ndvi)
```

### 2020-08-10 text/ecography/round_3/title_page.Rmd
migrate gitlab repo to github

Though tropical dry forests (TDF) are the most endangered tropical ecosystem, they are still understudied.
Therefore, we studied the spatio-temporal effects of different El Niño Southern Oscillation (ENSO) episodes on TDF biodiversity in the terrestrial core region of ENSO in NW Peru.
For this, we sampled 50 permanent vegetation plots along a climatic gradient in four years representing different ENSO episodes: dry La Niña (2011), humid La Niña (2012), moderate El Niño (2016) and a very humid neutral year (coastal El Niño; 2017).
To visually represent the spatio-temporal changes in floristic composition, we used an innovative approach by modeling the ordination scores using a Generalized Additive Model and spatial cross-validation.
Additionally, we applied variation partitioning to test if topographic or edaphic variables gained in importance as soon as water was no longer the limiting factor.
Plant species richness increased under humid conditions during the humid La Niña (2012) and moderate El Niño (2016) year, and slightly decreased under the very humid conditions of the coastal El Niño year (2017).
Vegetation composition changed also under more humid conditions.
Vegetation cover was manifold higher in 2017 compared to all previous years.
Edaphic and topographic variables gained in importance with increased water availability in 2012 and 2016, however, this effect was not further amplified under very wet conditions (2017).
As nutrients can be equally important for primary production as water, we additionally executed an irrigation-nutrient experiment.
The experiment showed that biomass production under wet Super Niño conditions was three times higher when a fertilizer treatment was added.
Overall, our results can help to 1) protect local human populations from devastating ENSO effects through informed conservation management and the restoration of degraded ecosystems; 2) equip farmers with information regarding sustainable agrarian management.

@@

*Funding* - We are also most thankful to the Deutsche Forschungsgemeinschaft (project Ri 370/19-1) and a Friedrich Schiller University of Jena program to support junior researchers (project DRM/2017-05) for funding this work.

*Author contributions* - JM, MR and AJ conceived the ideas.
JM, TB, DJ, RAR and PSZ conducted the ﬁeldwork and collected the data with the help of collaborators.
JM, PD, AB, TF, JB, GD, RR, DJ, PS, DN and EFR analyzed the data.
JM, DN and PS were responsible for establishing the research compendium.
JM led the writing with the assistance of all co-authors.

>>>>

The tropical dry forests of NW Peru are heavily shaped by the El Niño Southern Oscillation (ENSO), where especially El Niño brings rain to arid to semi-arid areas.
However, the resulting effects on biodiversity patterns remain largely unknown as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Therefore, we studied the spatio-temporal effects of different ENSO episodes on floristic biodiversity along a climatic gradient ranging from the coastal desert to the Andean foothills.
We sampled 50 vegetation plots in four years representing different ENSO episodes.
To highlight the spatio-temporal changes in floristic composition and beta diversity across ENSO episodes, we predicted ordination scores with a Generalized Additive Model.
We applied variation partitioning to test if topographic or edaphic variables gained in importance during more humid ENSO episodes.
Additionally, we executed an irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction under different simulated ENSO rainfall scenarios. 
Plant species richness increased under humid conditions during the humid La Niña (2012) and the moderate El Niño (2016), and slightly decreased under the very humid conditions during the coastal El Niño (2017).
The spatial prediction revealed that specific vegetation formations became more pronounced with increasing water input, but that a large water surplus led to the disruption of the strict order along the climatic gradient.
Edaphic and topographic variables gained in importance with increased water availability (2012 & 2016), however, this effect was not further amplified under very wet conditions (2017).
The experiment showed that plant cover under Super Niño conditions was three times higher when fertilized.
Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.

<!--
Overall, our results can help to 1) protect local human populations from devastating ENSO effects through informed conservation management and the restoration of degraded ecosystems; 2) equip farmers with information regarding sustainable agrarian management.
-->

@@

*Funding* - We are also most thankful to the Deutsche Forschungsgemeinschaft (project Ri 370/19-1), a Friedrich Schiller University of Jena program to support junior researchers (project DRM/2017-05) and the project "Ecología Funcional de los Bosques Andaluces y Predicciones Sobre Sus Cambios Futuros (For-Change)" (UCO-27943) by Junta de Andalucía (Spain) and European FEDER funds for funding this work.

*Author contributions* - JM, MR and AJ conceived the ideas.
JM, TB, DJ, RAR and PSZ conducted the ﬁeld work and collected the data with the help of collaborators.
JM, PD, AB, TF, JB, GD, RR, DJ, PS, DN and EFR analyzed the data.
JM, DN and PS were responsible for establishing the research compendium.
JM led the writing with the assistance of all co-authors.
